Quantifying the importance of an inference model in Bayesian phylogenetics

### 2020-05-27 pirouette_article.tex
Progress response letter 2

  \textbf{1. }
    Phylogenetic trees are currently routinely reconstructed from an alignment 
    of character sequences (usually nucleotide sequences). 
    Bayesian tools, such as MrBayes, RevBayes and BEAST2, 
    have gained much popularity over the last decade, 
    as they allow joint estimation of the posterior distribution of the 
    phylogenetic trees and the parameters of the underlying inference model.  
    An important ingredient of these Bayesian approaches is the species tree 
    prior.
    In principle, the Bayesian framework allows for comparing 
    different tree priors, which may elucidate the macroevolutionary 
    processes underlying the species tree.
    In practice, however, only macroevolutionary models that allow for fast computation of the prior probability are used.
    The question is how accurate the tree estimation is when the real macroevolutionary processes are substantially different from those assumed in the tree prior. \\
    \textbf{2. }
    Here we present \verb;pirouette;, 
    a free, libre and open-source R package that assesses 
    the inference error made by Bayesian phylogenetics for a given macroevolutionary diversification model. \verb;pirouette; makes use of 
    BEAST2, but its philosophy applies to any Bayesian phylogenetic inference 
    tool. \\
  \textbf{3. }
    We describe \verb;pirouette;'s usage and the biological scientific
    question it can answer, including full examples. \\
  \textbf{4. }
    Last, we discuss the results obtained by the examples and their 
    interpretation. \\
\end{abstract}

>>>>

  \textbf{1. }
    Phylogenetic trees are currently routinely reconstructed from an alignment 
    of character sequences (usually nucleotide sequences). 
    Bayesian tools, such as MrBayes, RevBayes and BEAST2, 
    have gained much popularity over the last decade, 
    as they allow joint estimation of the posterior distribution of the 
    phylogenetic trees and the parameters of the underlying inference model.  
    An important ingredient of these Bayesian approaches is the species tree 
    prior.
    In principle, the Bayesian framework allows for comparing 
    different tree priors, which may elucidate the macroevolutionary 
    processes underlying the species tree.
    In practice, however, only macroevolutionary models that allow for fast computation of the prior probability are used.
    The question is how accurate the tree estimation is when the real macroevolutionary processes are substantially different from those assumed in the tree prior. \\
    \textbf{2. }
    Here we present \verb;pirouette;, 
    a free and open-source R package that assesses 
    the inference error made by Bayesian phylogenetics for a given macroevolutionary diversification model. \verb;pirouette; makes use of 
    BEAST2, but its philosophy applies to any Bayesian phylogenetic inference 
    tool. \\
  \textbf{3. }
    We describe \verb;pirouette;'s usage and the biological scientific
    question it can answer, including full examples. \\
  \textbf{4. }
    Last, we discuss the results obtained by the examples and their 
    interpretation. \\
\end{abstract}

### 2020-05-27 pirouette_supplement.tex
Only one bib style and source
pirouette has refs to citations and figs
Stick to 80 chars per line
Progress response letter 2

This supplementary material contains additional facets of \verb;pirouette;, such as the installation of the package, an overview of
pirouette's main functions and a guide for empiricists, based on multiple experiments
that are shown here as well.

@@

\begin{itemize}
  \item{
    subsection \ref{subsec:guidelines}: guidelines for empiricists
  }
  \item{
    subsection \ref{subsec:installation}: installation
  }
  \item{
    subsection \ref{subsec:resources}: resources, such as 
    website, tutorials, packages used, bug reporting and contributing
  }
  \item{
    subsection \ref{subsec:citation}: citation of pirouette
  }
  \item{
    subsection \ref{subsec:main_functions}: main functions
  }
  \item{
    subsection \ref{subsec:main_example}: code, extra figures and
    diagnostics regarding the main example.
  }
  \item{
    subsection \ref{subsec:distribution}: the result of using 
    multiple trees, as generated by the
    same stochastic process as the main example
  }
  \item{
    subsection \ref{subsec:n_taxa}: the effect of the number
    of taxa
  }
  \item{
    subsection \ref{subsec:n_nucleotides}: the effect of the DNA
    alignment sequence length
  }
  \item{
    subsection \ref{subsec:simplest_correct_parameterization} shows the
    effect when doing an inference in the simplest use case
  }
  \item{
    subsection \ref{subsec:under_parameterization} shows the
    effect when doing an inference with an under-parameterization
  }
  \item{
    subsection \ref{subsec:different_n_mutations} shows the
    effect when the twin alignment is allowed to have a different
    number of mutations
  }
  \item{
    subsection \ref{subsec:mutation_rate} shows the
    effect of different mutation rates
  }
\end{itemize}

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\subsection{Guidelines for empiricists}
\label{subsec:guidelines}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

@@

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\begin{figure}[H]
  \centering
  \resizebox {1.0\columnwidth} {!} {
    \begin{tikzpicture}[
      ->,>=stealth',shorten >=1pt,auto,
      node distance=0.5\textheight, 
      semithick
    ]   
    \tikzstyle{every state}=[]
    \node[state, draw=none] (O) [] {
    };   
    \node[state] (A) [right of = O, rectangle] {
      \includegraphics[height=0.4\textheight]{pirouette_example_30/example_30_314/true_tree.png}
    };   
    \node[state] (B) [below of = A, rectangle] {
      \includegraphics[height=0.25\textheight]{pirouette_example_30/example_30_314/true_alignment.png}
    };   
    \node[state] (CG) [below of = B, rectangle] {
      \includegraphics[height=0.3\textheight]{pirouette_example_30/example_30_314/true_posterior_gen.png}
    };   
    \node[state] (DG) [below of = CG, rectangle] {
      \includegraphics[height=0.3\textheight]{pirouette_example_30/example_30_314/true_error_violin_gen.png}
    };   
    \node[state] (CB) [right of = CG, rectangle] {
      \includegraphics[height=0.3\textheight]{pirouette_example_30/example_30_314/true_posterior_best.png}
    };   
    \node[state] (DB) [below of = CB, rectangle] {
      \includegraphics[height=0.3\textheight]{pirouette_example_30/example_30_314/true_error_violin_best.png}
    };   
    \node[state] (AT) [right of = A, rectangle, node distance=0.8\textheight] {
      \includegraphics[height=0.4\textheight]{pirouette_example_30/example_30_314/twin_tree.png}
    };   
    \node[state] (BT) [below of = AT, rectangle] {
      \includegraphics[height=0.25\textheight]{pirouette_example_30/example_30_314/twin_alignment.png}
    };   
    \node[state] (CTG) [right of = CB, rectangle] {
      \includegraphics[height=0.3\textheight]{pirouette_example_30/example_30_314/twin_posterior_gen.png}
    };   
    \node[state] (DTG) [below of = CTG, rectangle] {
      \includegraphics[height=0.3\textheight]{pirouette_example_30/example_30_314/twin_error_violin_gen.png}
    };   
    \node[state] (CTB) [right of = CTG, rectangle] {
      \includegraphics[height=0.3\textheight]{pirouette_example_30/example_30_314/twin_posterior_best.png}
    };   
    \node[state] (DTB) [below of = CTB, rectangle] {
      \includegraphics[height=0.3\textheight]{pirouette_example_30/example_30_314/twin_error_violin_best.png}
    };   
    \path 
      (O) edge [anchor = south] node {} (A)
      (A) edge [anchor = south] node {} (B)
      (B) edge [anchor = south] node {} (CG)
      (CG) edge [anchor = south] node {} (DG)
      (B) edge [anchor = south east] node {} (CB)
      (CB) edge [anchor = south] node {} (DB)
      (A) edge [anchor = east] node {} (AT)
      (AT) edge [anchor = south] node {} (BT)
      (BT) edge [anchor = south east] node {} (CTG)
      (CTG) edge [anchor = south] node {} (DTG)
      (BT) edge [anchor = south] node {} (CTB)
      (CTB) edge [anchor = south] node {} (DTB)
    ; 
    \end{tikzpicture}
  }
  \caption{Full pirouette pipeline, including comparison to baseline error. The true tree (top right) is used to simulate an alignment. From this alignment two posterior distributions of trees are created: one using the generative model and another one using the inference model with the highest marginal likelihood. For each distribution of trees, a distribution of errors, measured with the nLTT statistic, between the posterior trees and the main trees is drawn. From the true tree also a twin tree is created (right side of the figure) which follows the same pipeline, leading to two additional error distributions to use as baseline errors.}
  \label{fig:example_30_full_pipeline}
\end{figure}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

@@

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
% Bibliography
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
% MEE style
\bibliographystyle{pirouette_mee}
\bibliography{pirouette_supplement}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

>>>>

This supplementary material contains additional facets of \verb;pirouette;, such as the installation of the package, an overview of
pirouette's main functions and a guide for users, based on multiple experiments
that are shown here as well.

@@

\begin{itemize}
  \item{
    subsection \ref{subsec:guidelines}: guidelines for users
  }
  \item{
    subsection \ref{subsec:installation}: installation
  }
  \item{
    subsection \ref{subsec:resources}: resources, such as 
    website, tutorials, packages used, bug reporting and contributing
  }
  \item{
    subsection \ref{subsec:citation}: citation of pirouette
  }
  \item{
    subsection \ref{subsec:main_functions}: main functions
  }
  \item{
    subsection \ref{subsec:main_example}: code, extra figures and
    diagnostics regarding the main example.
  }
  \item{
    subsection \ref{subsec:distribution}: the result of using 
    multiple trees, as generated by the
    same stochastic process as the main example
  }
  \item{
    subsection \ref{subsec:n_taxa}: the effect of the number
    of taxa
  }
  \item{
    subsection \ref{subsec:n_nucleotides}: the effect of the DNA
    alignment sequence length
  }
  \item{
    subsection \ref{subsec:simplest_correct_parameterization} shows the
    effect when doing an inference in the simplest use case
  }
  \item{
    subsection \ref{subsec:under_parameterization} shows the
    effect when doing an inference with an under-parameterization
  }
  \item{
    subsection \ref{subsec:different_n_mutations} shows the
    effect when the twin alignment is allowed to have a different
    number of mutations
  }
  \item{
    subsection \ref{subsec:mutation_rate} shows the
    effect of different mutation rates
  }
\end{itemize}

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\subsection{Guidelines for users}
\label{subsec:guidelines}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

@@

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\begin{figure}[H]
  \centering
  \resizebox {1.0\columnwidth} {!} {
    \begin{tikzpicture}[
      ->,>=stealth',shorten >=1pt,auto,
      node distance=0.5\textheight, 
      semithick
    ]   
    \tikzstyle{every state}=[]
    \node[state, draw=none] (O) [] {
    };   
    \node[state] (A) [right of = O, rectangle] {
      \includegraphics[height=0.4\textheight]{pirouette_example_30/example_30_314/true_tree.png}
    };   
    \node[state] (B) [below of = A, rectangle] {
      \includegraphics[height=0.25\textheight]{pirouette_example_30/example_30_314/true_alignment.png}
    };   
    \node[state] (CG) [below of = B, rectangle] {
      \includegraphics[height=0.3\textheight]{pirouette_example_30/example_30_314/true_posterior_gen.png}
    };   
    \node[state] (DG) [below of = CG, rectangle] {
      \includegraphics[height=0.3\textheight]{pirouette_example_30/example_30_314/true_error_violin_gen.png}
    };   
    \node[state] (CB) [right of = CG, rectangle] {
      \includegraphics[height=0.3\textheight]{pirouette_example_30/example_30_314/true_posterior_best.png}
    };   
    \node[state] (DB) [below of = CB, rectangle] {
      \includegraphics[height=0.3\textheight]{pirouette_example_30/example_30_314/true_error_violin_best.png}
    };   
    \node[state] (AT) [right of = A, rectangle, node distance=0.8\textheight] {
      \includegraphics[height=0.4\textheight]{pirouette_example_30/example_30_314/twin_tree.png}
    };   
    \node[state] (BT) [below of = AT, rectangle] {
      \includegraphics[height=0.25\textheight]{pirouette_example_30/example_30_314/twin_alignment.png}
    };   
    \node[state] (CTG) [right of = CB, rectangle] {
      \includegraphics[height=0.3\textheight]{pirouette_example_30/example_30_314/twin_posterior_gen.png}
    };   
    \node[state] (DTG) [below of = CTG, rectangle] {
      \includegraphics[height=0.3\textheight]{pirouette_example_30/example_30_314/twin_error_violin_gen.png}
    };   
    \node[state] (CTB) [right of = CTG, rectangle] {
      \includegraphics[height=0.3\textheight]{pirouette_example_30/example_30_314/twin_posterior_best.png}
    };   
    \node[state] (DTB) [below of = CTB, rectangle] {
      \includegraphics[height=0.3\textheight]{pirouette_example_30/example_30_314/twin_error_violin_best.png}
    };   
    \path 
      (O) edge [anchor = south] node {} (A)
      (A) edge [anchor = south] node {} (B)
      (B) edge [anchor = south] node {} (CG)
      (CG) edge [anchor = south] node {} (DG)
      (B) edge [anchor = south east] node {} (CB)
      (CB) edge [anchor = south] node {} (DB)
      (A) edge [anchor = east] node {} (AT)
      (AT) edge [anchor = south] node {} (BT)
      (BT) edge [anchor = south east] node {} (CTG)
      (CTG) edge [anchor = south] node {} (DTG)
      (BT) edge [anchor = south] node {} (CTB)
      (CTB) edge [anchor = south] node {} (DTB)
    ; 
    \end{tikzpicture}
  }
  \caption{
    Full pirouette pipeline, including comparison to baseline error. 
    The true tree (top left) is used to simulate an alignment. 
    From this alignment two posterior distributions of trees are created: 
    one using the generative model and another one using the inference model with the highest marginal likelihood. For each distribution of trees, a distribution of errors, measured with the nLTT statistic, between the posterior trees and the main trees is drawn. From the true tree also a twin tree is created (right side of the figure) which follows the same pipeline, leading to two additional error distributions to use as baseline errors.
  }
  \label{fig:example_30_full_pipeline}
\end{figure}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

@@

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
% Bibliography
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
% MEE style
\bibliographystyle{pirouette_mee}
% \bibliography{pirouette_supplement}
\bibliography{pirouette_article}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
